Breakpoint Models Show No Evidence of Recreational Adaptation to Increasing Wildfire Smoke in the American West

### 2021-07-07 MarkdownDocuments/Manuscript_Real.Rmd
Finished first draft of the text. Next up is fig captions

---
title: Breakpoint Models Show No Evidence of Recreational Adaptation to Increasing Wildfire Smoke in the American West
author:
- affiliation: Boise State University - Human Environment Systems
  name: Matt Clark
- affiliation: University of Michigan
  name: Alexander Killion
- affiliation: Boise State University
  name: Matthew A. Williamson
- affiliation: Boise State University
  name: Vicken Hillis
- affiliation: Boise State University
  name: Laura? 
date: "`r format(Sys.time(), '%B %d, %Y')`"
keywords: Wildfire, Smoke, Recreation, Park Science
bibliography: [../SmokePaper.bib, ../OtherRefs.bib]
abstract: Humans are going to have to adapt to our increasingly variable and hazardous climate. Rapidly increasing wildfire smoke in the American West represents a serious risk to human health and also a pertinent test-case for individual climate adaptation generally. Here we aggregate data on ambient smoke and recreational visitation in 32 U.S. national parks from 1980 - 2019 to test for observable adaptation to increasing wildfire smoke. We develop a series of breakpoint models to allow for threshold effects in recreational response. These models demonstrate that even when dangerous levels of smoke are present, individuals are failing to limit outdoor recreation. Future research may further our understanding of individual-level climate adaptation by focusing on alternative hazards or finer time series.

@@

```{r plotzsrc, include=FALSE, cache=TRUE, message = FALSE, warning=FALSE}
source("~/SmokeProject/Rcode/PaperPlotsFinal.R")
#source("~/SmokeProject/Rcode/PosteriorPlots.r")

@@

# Introduction
Climate change has already begun to dramatically change the social-ecological landscapes in which humans have learned to operate [@IPCC]. All levels of human society must urgently adapt to these new realities if we are to continue to thrive over the coming millenia [@IPCC]. As a society, we are aware of this, but still are failing to respond adequately [@steg_limiting_2018]. Even when adaptive strategies and their benefits are known, we do not observe adaptive behaviors [@van_valkengoed_meta-analyses_2019]. While research has demonstrated that extreme events tend to spur climate adaptation at the government level, less is known about how individuals are reacting and possibly adapting to a changing climate [@bierbaum_comprehensive_2013; @berrang-ford_tracking_2019; @berrang-ford_are_2011].

@@

# Methods
## Data Collection

### Visitation Data
Visitation data were obtained on the monthly scale through the [National Park Service Visitor Use Statistics Portal](https://irma.nps.gov/STATS/). These data are generally collected via car counters, permit information, and concessionaire reporting. Methods of collection vary from park to park. Unit specific information can be found on the National Park Service Visitor Use Statistics Portal. Although collection methods are inconsistent across the sample, all data was taken as reported. We retrieved monthly data for all national park units in the Intermountain and Pacific West regions for years 1980 through 2019. To reduce issues regarding park seasonality, we then subset our data to only the three month high season associated with each park. For example, only Summer (June, July, Aug) data was used for Yellowstone National Park.

### SmokeData
Alex to fill in this part

@@

# Discussion
We demonstrate that given the data available to us, we detect no visitor adaptation to increasing wildfire smoke in national parks in the American West. This result is troubling both immediately and in the long-term. We discuss earlier the public health dangers of wildfire smoke, and in particular $\text{PM}_{2.5}$. These results suggest that..

We probably need national level policy to protect people [@rai_policy_2020]. There is huge variation in the way states react and plan for climate change [@hu_evaluating_2018].

because of the highly variable nature of wildfires, their relationship to climate change, and the great distances that smoke can travel during and after wildfire events [@dombeck_wildfire_2004; @collins_regional_2006].

A key limitation of this study is the lack of availability of fine scale visitation data to US national parks. Daily visitation data would allow us to identify a more complete picture of if and how visitors respond to smoke events day to day. Current data at the monthly scale allows us to investigate the big picture trends in visitor behavior, but future research would benefit greatly from daily count data.

It is possible that individuals postpone visits to national parks by days or weeks when smoke is present, behavior which our study current study would be unable to capture. It is also possibly that this trend would not hold for recreation on all public lands. US national parks draw a great number non-local visitors, many of whom are coming from other states or countries and are likely visiting for the first time [@schuett2010visits]. We speculate that these visitors are less likely to change their plans due to wildfire smoke than individuals recreating locally or repeatedly in one location. We propose that this unique feature of national parks makes their visitors particularly less adaptive to changing climatic conditions. Another path for future research might investigate if the trend shown here holds for other public lands with different visitor profiles.

>>>>

---
title: Breakpoint Models Show No Evidence of Recreational Adaptation to Increasing Wildfire Smoke in the American West
author:
- affiliation: Boise State University - Human Environment Systems
  name: Matt Clark
- affiliation: University of Michigan
  name: Alexander Killion
- affiliation: Boise State University
  name: Matthew A. Williamson
- affiliation: Boise State University
  name: Vicken Hillis
- affiliation: Boise State University
  name: Laura? 
date: "`r format(Sys.time(), '%B %d, %Y')`"
keywords: Wildfire, Smoke, Recreation, Park Science
bibliography: [../SmokePaper.bib, ../OtherRefs.bib]
abstract: Humans must adapt to our increasingly variable and hazardous climate. Worsening wildfire smoke events in the American West represent a serious risk to human health and also a pertinent test-case for individual climate adaptation generally. Here we aggregate data on ambient smoke and recreational visitation in 32 U.S. national parks from 1980 - 2019 to test for observable adaptation to worsening wildfire smoke. We develop a series of breakpoint models to allow for threshold effects in recreational response. These models demonstrate that even when dangerous levels of smoke are present, individuals are failing to limit outdoor recreation. Future research may further our understanding of individual-level climate adaptation by focusing on alternative hazards or finer time series.

@@

```{r plotzsrc, include=FALSE, cache=TRUE, message = FALSE, warning=FALSE}
#source("~/SmokeProject/Rcode/PaperPlotsFinal.R")
#source("~/SmokeProject/Rcode/PosteriorPlots.r")

@@

# Introduction
Climate change has already begun to dramatically change the social-ecological landscapes in which humans have learned to operate [@IPCC]. All levels of human society must urgently adapt to these new realities if we are to continue to thrive over the coming millenia [@IPCC]. As a society, we are aware of this, but still are failing to respond adequately [@steg_limiting_2018]. Even when adaptive strategies and their benefits are known, we do not demonstrate adaptive behaviors [@van_valkengoed_meta-analyses_2019]. While research has shown that extreme events tend to spur climate adaptation at the government level, less is known about how individuals are reacting and possibly adapting to a changing climate [@bierbaum_comprehensive_2013; @berrang-ford_tracking_2019; @berrang-ford_are_2011].

@@

# Methods
## Data collection

### Visitation data
Visitation data were obtained on the monthly scale through the [National Park Service Visitor Use Statistics Portal](https://irma.nps.gov/STATS/). These data are generally collected via car counters, permit information, and concessionaire reporting. Methods of collection vary from park to park. Unit specific information can be found on the National Park Service Visitor Use Statistics Portal. Although collection methods are inconsistent across the sample, all data was taken as reported. We retrieved monthly data for all national park units in the Intermountain and Pacific West regions for years 1980 through 2019. To reduce issues regarding park seasonality, we then subset our data to only the three month high season associated with each park. For example, only Summer (June, July, Aug) data was used for Yellowstone National Park.

### Smoke data
Alex to fill in this part

@@

# Discussion
Given the data available to us, we detect no visitor adaptation to increasing wildfire smoke in national parks in the American West. This result is troubling both immediately and in the long-term. As discussed above, $\text{PM}_{2.5}$ significantly increases community morbidity even under acute exposure. In concert with showing no overall trends in behavioral adaptation, these data provide specific instances of historic smoke events where visitation did not deviate from normal.

The highly variable nature of wildfires, their relationship to climate change, and the great distances that smoke can travel during and after wildfire events make it difficult for individuals to plan around smoke events [@dombeck_wildfire_2004; @collins_regional_2006]. In conjunction, U.S. national parks draw a great number of non-local visitors, many of whom are coming from other states or countries and are likely visiting for the first time [@schuett2010visits]. We speculate that these visitors are less likely to change their plans due to wildfire smoke than individuals recreating locally or repeatedly in one location. We propose that these unique features of national parks and wildfire smoke make visitors particularly unable to adapt to changing climatic conditions.

Based on our findings here, we put forward that a regional or national level policy limiting visitation during dangerous smoke events may be necessary to protect would-be visitors to U.S. national parks [@rai_policy_2020]. Presently, there is considerable variation in the way states react and plan for climate change and associated hazards [@hu_evaluating_2018].

## Limitations and fututre research

A key limitation of this study is the lack of availability of fine scale visitation data to US national parks. Daily visitation data would allow us to identify a more complete picture of if and how visitors respond to smoke events day to day. Current data at the monthly scale allows us to investigate the big picture trends in visitor behavior, but future research would benefit greatly from daily count data. It is possible that individuals postpone visits to national parks by days or weeks when smoke is present, behavior which our current study would be unable to capture.

It is also possible that this trend would not be observed for recreation on all public lands or for all climate change-induced hazards. Another path for future research might investigate if the trend shown here holds for other public lands with different visitor profiles and for other hazards.

# Conclusion 
The results presented here indicate that individuals are not modifying their behavior to adapt to worsening wildfire smoke events. It is unclear if this finding is representative of individual climate adaptation generally or is unique to this system. Regardless of the generality of our findings, this study has specific implications for management of U.S. national parks. Future research may confirm or negate the trend shown here in individual climate adaptation broadly. This will have considerable implications for the future of all communities living with climate-induced hazards.
